\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-04-22 paper.qmd
Changed the scale
re-working hexagon grid computation

With our goal being to make a representation of this 2D layout that can be lifted into high-dimensional space, the layout needs to be augmented to include neighbour information. A simple approach would be to triangulate the points and add edges. A more stable approach is to first hexagonally bin the data, and connect the bin centroids. This process serves to reduce some noisiness in the resulting surface shown in \pD{}. The steps in this process are shown in @fig-NLDR-scurve, and documented below.

@@

## Hexagonal bining to have regular hexagons
hb_obj_scurve <- hex_binning(
  data = umap_scurve_scaled, 
  x = "UMAP1",
  y = "UMAP2", 
  num_bins_x = num_bins_x_scurve,
  num_bins_y = num_bins_y_scurve, 
  x_start = x_start_scurve,
  y_start = y_start_scurve, 
  buffer_x = buffer_x_scurve,
  buffer_y = buffer_y_scurve, 
  hex_size = hex_size_scurve,
  col_start = "UMAP")

@@

It is beneficial to define the algorithm on data having a standard scale. Here we use [0, 1] scaling, but by default include the aspect ratio produced by the NLDR $(r_1, r_2, ..., r_k)$, by allowing for the upper bound to vary, $(y_{1,\text{max}}, y_{2,\text{max}}, ..., y_{k,\text{max}})$.

When $k=2$ which is assumed for hexagon binning, $y_{1,\text{max}}=1$ and $y_{2,\text{max}} = \frac{r_2}{r_1}\times \frac{2}{\sqrt{3}}$. (The $\frac{2}{\sqrt{3}}$ accounts for the different height ($a_1$) and width ($a_2$) of a regular hexagon.) (@fig-NLDR-scurve (a))

#### Computing hexagonal grid configurations

To define a hexagonal grid across the \kD{} space, with hexagons of fixed height ($a_1$), width ($a_2$),  it is necessary to determine how many hexagons should be represented along each axis in the \kD{} space.

When $k=2$, the number of bins along the $x$ and $y$ axes, $b_1$ and $b_2$, is computed by considering the scaling factors ${r_1, r_2}$, along with the offset along the axes ($q_1$, $q_2$), and the height ($a_1$) and width ($a_2$) of a regular hexagon (@fig-NLDR-scurve (b)). ($b_1 = \frac{r_1 + q_1}{a_2}$, and $b_2 = \frac{r_2 + q_2}{0.75 \times a_1}$) where $0.75 \times a_1$ accommodate to have hexagons fill the entire 2D space without leaving any gaps between them.)

>>>>

With our goal being to make a representation of this 2D layout that can be lifted into high-dimensional space, the layout needs to be augmented to include neighbour information. A simple approach would be to triangulate the points and add edges. A more stable approach is to first hexagonally bin the data, reducing it from $n$ to $m\leq n$ observations, and connect the bin centroids. This process serves to reduce some noisiness in the resulting surface shown in \pD{}. The steps in this process are shown in @fig-NLDR-scurve, and documented below.

@@

## Hexagonal binning to have regular hexagons
hb_obj_scurve <- hex_binning(
  data = umap_scurve_scaled, 
  x = "UMAP1",
  y = "UMAP2", 
  num_bins_x = num_bins_x_scurve,
  num_bins_y = num_bins_y_scurve, 
  x_start = x_start_scurve,
  y_start = y_start_scurve, 
  buffer_x = buffer_x_scurve,
  buffer_y = buffer_y_scurve, 
  hex_size = hex_size_scurve,
  col_start = "UMAP")

@@

It is beneficial to define the algorithm on data having a standard scale. Here the variables are scaled to [0, 1], but the upper bound can incorporate the aspect ratio produced by the NLDR $(r_1, r_2, ..., r_k)$, by setting them to $(y_{1,\text{max}}, y_{2,\text{max}}, ..., y_{k,\text{max}})$. When $k=2$ which is assumed for hexagon binning, $y_{1,\text{max}}=1$ and $y_{2,\text{max}} = \frac{r_2}{r_1}$, as observed in @fig-NLDR-scurve. <!-- \times \frac{2}{\sqrt{3}}$. (The $\frac{2}{\sqrt{3}}$ accounts for the different height ($a_1$) and width ($a_2$) of a regular hexagon.) The scaling of data should take the size of the hexagons into account, but choice of number of bins should. -->

#### Computing hexagon grid configurations

The 2D hexagon grid is defined by the number of bins in each direction ($b_1, b_2$), and the lower left position where the grid starts ($s_1, s_2$). The values of $s_i$ need to be below their respective minimum variable values, and could be a full bin lower, to allow a buffer corresponding to a full hexagon width and height around the data. The values of $b_i$ are variables to be computed that define the reduction in size of the data ($n$ to $m$).

Number of bins is set by fixing $b_1$, which determines the binwidth and breaks on $x_1$, is calculated as

The computed binwidth then determines the number of vertical bins, $b_2$, and vertical binwidth, which are computed based on $r_2$, and the height/width ratio of a regular hexagon, $\frac{2}{\sqrt{3}}$ as

<!-- To define a hexagonal grid across the \kD{} space, with hexagons of fixed height ($a_1$), width ($a_2$),  it is necessary to determine how many hexagons should be represented along each axis in the \kD{} space.

When $k=2$, the number of bins along the $x$ and $y$ axes, $b_1$ and $b_2$, is computed by considering the scaling factors ${r_1, r_2}$, along with the offset along the axes ($q_1$, $q_2$), and the height ($a_1$) and width ($a_2$) of a regular hexagon (@fig-NLDR-scurve (b)). ($b_1 = \frac{r_1 + q_1}{a_2}$, and $b_2 = \frac{r_2 + q_2}{0.75 \times a_1}$) where $0.75 \times a_1$ accommodate to have hexagons fill the entire 2D space without leaving any gaps between them.)-->

<!-- XXX We don't need parameters for the regular hexagon, these are fixed constants. -->

### 2024-04-22 paper.tex
re-working hexagon grid computation

With our goal being to make a representation of this 2D layout that can
be lifted into high-dimensional space, the layout needs to be augmented
to include neighbour information. A simple approach would be to
triangulate the points and add edges. A more stable approach is to first
hexagonally bin the data, and connect the bin centroids. This process
serves to reduce some noisiness in the resulting surface shown in \pD{}.
The steps in this process are shown in Figure~\ref{fig-NLDR-scurve}, and
documented below.

@@

It is beneficial to define the algorithm on data having a standard
scale. Here we use {[}0, 1{]} scaling, but by default include the aspect
ratio produced by the NLDR \((r_1, r_2, ..., r_k)\), by allowing for the
upper bound to vary,
\((y_{1,\text{max}}, y_{2,\text{max}}, ..., y_{k,\text{max}})\).

When \(k=2\) which is assumed for hexagon binning,
\(y_{1,\text{max}}=1\) and
\(y_{2,\text{max}} = \frac{r_2}{r_1}\times \frac{2}{\sqrt{3}}\). (The
\(\frac{2}{\sqrt{3}}\) accounts for the different height (\(a_1\)) and
width (\(a_2\)) of a regular hexagon.) (Figure~\ref{fig-NLDR-scurve}
(a))

\subsubsection{Computing hexagonal grid
configurations}\label{computing-hexagonal-grid-configurations}

To define a hexagonal grid across the \kD{} space, with hexagons of
fixed height (\(a_1\)), width (\(a_2\)), it is necessary to determine
how many hexagons should be represented along each axis in the \kD{}
space.

When \(k=2\), the number of bins along the \(x\) and \(y\) axes, \(b_1\)
and \(b_2\), is computed by considering the scaling factors
\({r_1, r_2}\), along with the offset along the axes (\(q_1\), \(q_2\)),
and the height (\(a_1\)) and width (\(a_2\)) of a regular hexagon
(Figure~\ref{fig-NLDR-scurve} (b)). (\(b_1 = \frac{r_1 + q_1}{a_2}\),
and \(b_2 = \frac{r_2 + q_2}{0.75 \times a_1}\)) where
\(0.75 \times a_1\) accommodate to have hexagons fill the entire 2D
space without leaving any gaps between them.)

>>>>

With our goal being to make a representation of this 2D layout that can
be lifted into high-dimensional space, the layout needs to be augmented
to include neighbour information. A simple approach would be to
triangulate the points and add edges. A more stable approach is to first
hexagonally bin the data, reducing it from \(n\) to \(m\leq n\)
observations, and connect the bin centroids. This process serves to
reduce some noisiness in the resulting surface shown in \pD{}. The steps
in this process are shown in Figure~\ref{fig-NLDR-scurve}, and
documented below.

@@

It is beneficial to define the algorithm on data having a standard
scale. Here the variables are scaled to {[}0, 1{]}, but the upper bound
can incorporate the aspect ratio produced by the NLDR
\((r_1, r_2, ..., r_k)\), by setting them to
\((y_{1,\text{max}}, y_{2,\text{max}}, ..., y_{k,\text{max}})\). When
\(k=2\) which is assumed for hexagon binning, \(y_{1,\text{max}}=1\) and
\(y_{2,\text{max}} = \frac{r_2}{r_1}\), as observed in
Figure~\ref{fig-NLDR-scurve}.

\subsubsection{Computing hexagon grid
configurations}\label{computing-hexagon-grid-configurations}

The 2D hexagon grid is defined by the number of bins in each direction
(\(b_1, b_2\)), and the lower left position where the grid starts
(\(s_1, s_2\)). The values of \(s_i\) need to be below their respective
minimum variable values, and could be a full bin lower, to allow a
buffer corresponding to a full hexagon width and height around the data.
The values of \(b_i\) are variables to be computed that define the
reduction in size of the data (\(n\) to \(m\)).

Number of bins is set by fixing \(b_1\), which determines the binwidth
and breaks on \(x_1\), is calculated as

The computed binwidth then determines the number of vertical bins,
\(b_2\), and vertical binwidth, which are computed based on \(r_2\), and
the height/width ratio of a regular hexagon, \(\frac{2}{\sqrt{3}}\) as
